---
title: "SAEH Model Building"
subtitle: "Purpose: Run longitunidal model and model fit diagnostics"
subject: "PI: Emily Boniface"
author: "Kelsey MacKenzie"
date: February 5, 2026
format: html
editor: visual
---

```{r, include = FALSE}
##### Load packages 
library(dplyr)
library(glmmTMB)
library(nlme)
library(lme4)
library(ggplot2)
library(effects)
library(geepack)
library(doBy)
library(emmeans)
library(gt)
library(tidyverse)
library(haven)

#### Import Data
SAEH_model_data <- readRDS("SAEH_dataset.rds")
```

# Exploration

```{r}
# Looking at variables of interest
table(SAEH_model_data$gestat, SAEH_model_data$marg_group, useNA = "always")

# How many records exceed 43 weeks for gestational duration?
SAEH_model_data %>%
  filter(gestat > 43) %>%
  count(gestat, sort = TRUE)

### Specifically the 88s
sum(SAEH_model_data$gestat == 88, na.rm = TRUE)

### Recode only 88 to NA (coded as "Not applicable" in the dictionary)
SAEH_model_data <- SAEH_model_data %>%
  mutate(gestat = ifelse(gestat == 88, NA, gestat))

### Verify that fix
summary(SAEH_model_data$gestat)

# Make sure gestation is numeric
SAEH_model_data <- SAEH_model_data %>% mutate(gestat_clean = as.numeric(gestat))

# Binary: Late (≥13 weeks, 2nd trimester+) vs Early (<13 weeks)
SAEH_full <- SAEH_model_data %>%
  mutate(gest_age_cat = case_when(
    gestat_clean <= 12 ~ "Early",
    gestat_clean >= 13 ~ "Late",
    is.na(gestat_clean) ~ NA_character_)) %>%
  mutate(gest_age_cat = factor(gest_age_cat, levels = c("Early", "Late")))

### Distribution of the cleaned continuous variable
summary(SAEH_full$gestat_clean)

### Distribution of the categorized variable
table(SAEH_full$gest_age_cat, useNA = "always")

### Cross-tab with marginalization group to get a first look
table(SAEH_full$gest_age_cat, SAEH_full$marg_group, useNA = "always")

# What percentage have gestat data? Split early/late?
prop.table(table(SAEH_full$gest_age_cat), margin = NULL) * 100  # %
```

## Plots

```{r}
# Plot 1: Mean gestational duration by year, split by marginalization group
p1 <- SAEH_full %>%
  filter(!is.na(gestat_clean) & !is.na(marg_group)) %>%
  group_by(year, marg_group) %>%
  summarise(mean_gest = mean(gestat_clean, na.rm = TRUE),
            se = sd(gestat_clean, na.rm = TRUE) / sqrt(n()),
            .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_gest, color = marg_group, group = marg_group)) +
  geom_point(size = 3) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_gest - 1.96*se,
                    ymax = mean_gest + 1.96*se), width = 0.2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  scale_color_manual(values = c("Less marginalized" = "steelblue",
                                "More marginalized" = "tomato")) +
  labs(title = "Mean gestational duration at presentation by year and marginalization group",
       x = "Year of discharge",
       y = "Mean gestational duration (weeks)",
       color = "Marginalization group") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10, hjust = 0.5))

p1

# Plot 2: Proportion of Late presentations by year and marginalization group
p2 <- SAEH_full %>%
  filter(!is.na(gest_age_cat) & !is.na(marg_group)) %>%
  group_by(year, marg_group) %>%
  summarise(prop_late = mean(gest_age_cat == "Late", na.rm = TRUE),
            .groups = "drop") %>%
  ggplot(aes(x = year, y = prop_late, color = marg_group, group = marg_group)) +
  geom_point(size = 3) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  scale_color_manual(values = c("Less marginalized" = "steelblue",
                                "More marginalized" = "tomato")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of late presentations by year and marginalization group",
       x = "Year of discharge",
       y = "Proportion with late presentation (≥13 weeks)",
       color = "Marginalization group") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

p2

# Save plots as PNG
ggsave("mean_gestational_duration_by_year.png", p1, width = 7, height = 5, dpi = 300)
ggsave("prop_late_by_year.png", p2, width = 7, height = 5, dpi = 300)
```

## Final Checks

```{r}
# Make year numeric and set reference group for marg_group
SAEH_full <- SAEH_full %>% 
  mutate(year_num = as.numeric(year), 
         marg_group = relevel(marg_group, ref = "Less marginalized"))

# Mean gest by year (quantify flatness)
SAEH_full %>%
  filter(!is.na(gestat_clean)) %>%
  group_by(year_num) %>%
  summarise(mean_gest = mean(gestat_clean), n = n(), .groups = "drop") %>%
  print(n = 7)

# Verify ALL Calkini (04001) records have grade_marg in SAEH_full --------------
cat("🔍 Calkini (04001) grade_marg verification:\n")
SAEH_full %>%
  filter(munic_id == "04001") %>%
  summarise(
    total_calkini_records = n(),
    missing_grade_marg = sum(is.na(grade_marg)),
    pct_complete = round(100 * mean(!is.na(grade_marg)), 1),
    unique_grade_marg = paste(unique(grade_marg[!is.na(grade_marg)]), collapse = ", "))

# Show sample of Calkini records
SAEH_full %>%
  filter(munic_id == "04001") %>%
  select(year, clues, munic_id, grade_marg) %>%
  print()


# FORCE all Calkini (04001) records to have grade_marg = "Bajo" ----------------
# Count NAs before fix
n_na_before <- sum(SAEH_full$munic_id == "04001" & is.na(SAEH_full$grade_marg), na.rm = TRUE)
cat("- Calkini NAs before:", n_na_before, "\n")

# FORCE "Bajo" for ALL 04001 records
SAEH_full <- SAEH_full %>%
  mutate(
    grade_marg = case_when(
      munic_id == "04001" ~ "Bajo",  # Force Calkini to Bajo
      TRUE ~ grade_marg))             # Keep others unchanged

# Verify 100% complete
cat("✓ After fix:\n")
SAEH_full %>%
  filter(munic_id == "04001") %>%
  summarise(
    total_calkini = n(),
    calkini_nas = sum(is.na(grade_marg)),
    all_bajo = sum(grade_marg == "Bajo", na.rm = TRUE),
    pct_complete = paste0(round(100 * mean(!is.na(grade_marg)), 1), "%")) %>%
  print()

# Save data for Emily
saveRDS(SAEH_full, "SAEH_1.rds")
names(SAEH_full) <- gsub("\\.", "_", names(SAEH_full)) # save as .dta
write_dta(SAEH_full, "SAEH_1.dta")

# And a quick look at how mean gestational duration varies by marginalization group and year
aggregate(gestat_clean ~ year + marg_group, data = SAEH_full, 
          FUN = function(x) round(mean(x, na.rm = TRUE), 2))

# Only complete cases - takes some observations due to missing gestational duration
SAEH_complete <- SAEH_full[complete.cases(SAEH_full[, c("gest_age_cat", "marg_group", "year", "id")]), ]
```

**MODEL EQUATION:**

$$\begin{aligned}
\text{logit}(P(Y_{ij}=1)) = \; &\beta_0 + \beta_1 \cdot \text{marg\_group}_{ij} + \beta_2 \cdot \text{year\_c}_{ij} + \beta_3 \cdot \text{year\_c}_{ij}^2 \; + \\
&\beta_4(\text{marg\_group}_{ij} \times \text{year\_c}_{ij}) + \beta_5(\text{marg\_group}_{ij} \times \text{year\_c}_{ij}^2) + u_j
\end{aligned}$$

**Where:**

- $\text{logit}(P(Y_{ij}=1))$ = log-odds of late presentation ($\geq$ 13 weeks gestational duration) for record $i$ in municipality $j$
- $\beta_0$ = baseline log-odds of late presentation for a Less marginalized municipality at the average year
- $\beta_1 \cdot \text{marg\_group}_{ij}$ = difference in log-odds for More vs. Less marginalized municipalities at the average year
- $\beta_2 \cdot \text{year\_c}_{ij}$ = linear component of the time trend for the Less marginalized reference group
- $\beta_3 \cdot \text{year\_c}_{ij}^2$ = quadratic component of the time trend
- $\beta_4(\text{marg\_group}_{ij} \times \text{year\_c}_{ij})$ = difference in the linear trajectory over time between marginalization groups
- $\beta_5(\text{marg\_group}_{ij} \times \text{year\_c}_{ij}^2)$ = difference in the curvature of the time trend between marginalization groups
- $u_j \sim \mathcal{N}(0, \sigma^2_u)$ = municipality-level random intercept

*Binary outcome, two groups, time trend that's non-linear, quadratic term, interaction between group and time*

## Prep for Model

```{r}
# Create centered year variable
mean_year <- mean(SAEH_complete$year, na.rm = TRUE)
mean_year  # year centers on ~ 2020 (2020.92)

# Now a year squared
SAEH_complete <- SAEH_complete %>%
  mutate(year_c  = year - mean_year,
         year_sq = year_c * year_c)

# Verify
SAEH_complete %>%
  select(year, year_c, year_sq) %>%
  distinct() %>%
  arrange(year)
```

# Prepare analytic dataset

```{r}
# Create numeric binary outcome (GEE needs 0/1)
SAEH_analytic <- SAEH_complete %>%
  mutate(late_pres = as.integer(gest_age_cat == "Late")) %>% # 1 = Late, 0 = Early
  arrange(munic_id, year)

# Confirm outcome coding
table(SAEH_analytic$late_pres, SAEH_analytic$gest_age_cat)

# Confirm analytic sample
nrow(SAEH_analytic)

# Verify the exact year_c value for 2021
knot_2021 <- SAEH_analytic %>%
  filter(year == 2021) %>%
  summarise(year_c_2021 = first(year_c)) %>%
  pull(year_c_2021)

# Build the spline term
SAEH_analytic <- SAEH_analytic %>%
  mutate(year_c_post = (year_c - knot_2021) * as.integer(year_c > knot_2021))

# Verify
SAEH_analytic %>%
  select(year, year_c, year_c_post) %>%
  distinct() %>%
  arrange(year)
```

## Handle covariates

```{r}
SAEH_analytic <- SAEH_analytic %>%
  mutate(
    # Convert parity to factor with 0 as reference, keep Missing as category
    parity_cat = factor(parity_cat, 
                        levels = c("0", "1", "2", "3+", "Missing")),
    
    # Age reference = 20-24 (already ordered correctly, just confirm)
    age_cat = relevel(as.factor(age_cat), ref = "20-24"),
    
    # Indigenous - keep No response as separate category
    indigena_bin = relevel(as.factor(indigena_bin), ref = "No"),
    
    # Region - North already reference
    region = relevel(as.factor(region), ref = "North"))

# Verify
levels(SAEH_analytic$parity_cat)
levels(SAEH_analytic$age_cat)
levels(SAEH_analytic$indigena_bin)
levels(SAEH_analytic$region)
```

# Making gee model work!

```{r}
# Aggregate municipality marginalization group as a fixed attribute
## Each municipality is measured 7 times (once per year)
## Use the dominant/assigned marg_group for each municipality
# SAEH_gee_agg <- SAEH_analytic %>%
#   group_by(munic_id, year, year_c, year_sq, year_c_post) %>%
#   summarise(n_total   = n(),
#             n_late    = sum(late_pres),
#             prop_late = mean(late_pres),
#             marg_group = first(marg_group),  # municipality-level characteristic
#             .groups   = "drop") %>%
#   arrange(munic_id, year)

# Check structure
# nrow(SAEH_gee_agg)
# head(SAEH_gee_agg)

# Check rows per municipality
# SAEH_gee_agg %>%
#   count(munic_id) %>%
#   count(n, name = "n_municipalities")
```

# Model

```{r}
# Logistic mixed effects model
model_glmer <- glmer(late_pres ~ marg_group * year_c + marg_group * year_sq + 
                       age_cat + parity_cat + indigena_bin + region + (1 | munic_id),
                     data    = SAEH_analytic,
                     family  = binomial(link = "logit"),
                     nAGQ    = 30,
                     control = glmerControl(optimizer = "bobyqa",
                                            optCtrl   = list(maxfun = 2e5)))


# GEE quadratic 
# model_gee <- geeglm(prop_late ~ marg_group * year_c + marg_group * year_sq,
#                     data    = SAEH_gee_agg,
#                     family  = gaussian(link = "identity"),
#                     weights = n_total,
#                     control = geese.control(epsilon = 0.0001, maxit = 100),
#                     id      = munic_id,
#                     corstr  = "exchangeable",
#                     std.err = "san.se")


# GEE linear spline (knot at 2021)
# model_spline <- geeglm(prop_late ~ marg_group * year_c + marg_group * year_c_post,
#                        data    = SAEH_gee_agg,
#                        family  = gaussian(link = "identity"),
#                        weights = n_total,
#                        control = geese.control(epsilon = 0.0001, maxit = 100),
#                        id      = munic_id,
#                        corstr  = "exchangeable",
#                        std.err = "san.se")


# Compare models
summary(model_glmer)
#summary(model_gee)
#summary(model_spline)
#QIC(model_gee)
#QIC(model_spline)
```

# Choosing final model

```{r}
# Fit comparison models
model_glmer_main <- glmer(late_pres ~ marg_group + year_c + year_sq + age_cat + 
                            parity_cat + indigena_bin + region + (1 | munic_id),
                          data    = SAEH_analytic,
                          family  = binomial(link = "logit"),
                          nAGQ    = 1,
                          control = glmerControl(optimizer = "bobyqa",
                                                 optCtrl   = list(maxfun = 2e5))) # main effects only (no interaction)

model_glmer_linear <- glmer(late_pres ~ marg_group * year_c +age_cat + parity_cat + 
                              indigena_bin + region + (1 | munic_id),
                            data    = SAEH_analytic,
                            family  = binomial(link = "logit"),
                            nAGQ    = 1,
                            control = glmerControl(optimizer = "bobyqa",
                                                   optCtrl   = list(maxfun = 2e5))) # linear interaction only (no quadratic)

# Compare
AIC(model_glmer_main, model_glmer_linear, model_glmer)
BIC(model_glmer_main, model_glmer_linear, model_glmer)
anova(model_glmer_main, model_glmer)
```

**Final model:** Mixed-effects logistic regression ('model_glmer')

Rationale: Data consists of individual hospital discharge records clustered within municipalities. No individual-level repeated measures structure exists. Municipality-level random intercept accounts for geographic clustering. GEE was not used as final model due to incompatibility between the data structure and GEE clustering requirements.

```{r}
# Odds ratios and 95% CIs for fixed effects
est <- esticon(model_glmer, diag(22))
est$Lower.CI <- est$estimate - 1.96 * est$std.error
est$Upper.CI <- est$estimate + 1.96 * est$std.error

OR.CI <- exp(cbind(est$estimate, est$Lower.CI, est$Upper.CI))
rownames(OR.CI) <- names(fixef(model_glmer))
colnames(OR.CI) <- c("OR", "Lower.OR", "Upper.OR")
round(OR.CI, 3)

# Also get the ICC to report
var_munic <- as.numeric(VarCorr(model_glmer)$munic_id)
icc <- var_munic / (var_munic + (pi^2 / 3))
round(icc, 3)
```

```{r}
# Extract coefficients
b0        <- fixef(model_glmer)["(Intercept)"]
b_marg    <- fixef(model_glmer)["marg_groupMore marginalized"]
b_yearc   <- fixef(model_glmer)["year_c"]
b_yearsq  <- fixef(model_glmer)["year_sq"]
b_int_c   <- fixef(model_glmer)["marg_groupMore marginalized:year_c"]
b_int_sq  <- fixef(model_glmer)["marg_groupMore marginalized:year_sq"]

# Year range on centered scale
year_range <- seq(min(SAEH_analytic$year_c), max(SAEH_analytic$year_c), by = 0.1)

# Predicted log-odds for each group
less_logits <- b0 + b_yearc * year_range + b_yearsq * year_range^2

more_logits <- b0 + b_marg +
               (b_yearc + b_int_c) * year_range +
               (b_yearsq + b_int_sq) * year_range^2

# Convert to probabilities
less_probs <- exp(less_logits) / (1 + exp(less_logits))
more_probs <- exp(more_logits) / (1 + exp(more_logits))

# Back-transform year_c to actual year for x-axis
actual_year <- year_range + mean_year

# Build plot data
plot_data <- data.frame(
  year  = rep(actual_year, 2),
  prob  = c(less_probs, more_probs),
  group = rep(c("Less marginalized", "More marginalized"), 
              each = length(actual_year)))

# Plot the estimated prob
ggplot(plot_data, aes(x = year, y = prob, color = group)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("Less marginalized" = "steelblue",
                                "More marginalized" = "tomato")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(
    title = "Estimated probability of late presentation by year and marginalization group",
    x     = "Year of discharge",
    y     = "Predicted probability of late presentation (≥13 weeks)",
    color = "Marginalization group") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

# Equivalent to margins command in Stata
mean_year <- mean(SAEH_analytic$year_num, na.rm = TRUE) # mean year value used for centering

# Convert target years to centered values
year_c_vals <- 2018:2024 - mean_year

emmeans(model_glmer, ~ marg_group * year_c,
        at   = list(year_c = year_c_vals),
        type = "response")
```

# Predicted prob table for presentation slides

```{r}
# Store emmeans output first
emmeans_out <- emmeans(model_glmer, ~ marg_group * year_c, 
                       at   = list(year_c = year_c_vals),
                       type = "response")

# Convert to dataframe and clean up
emmeans_df <- as.data.frame(emmeans_out) %>%
  mutate(year = round(year_c + mean_year))

# Create formatted estimate (95% CI) column
emmeans_df <- emmeans_df %>%
  mutate(est_ci = paste0(
    scales::percent(prob, accuracy = 0.1),
    " (", scales::percent(asymp.LCL, accuracy = 0.1),
    "–", scales::percent(asymp.UCL, accuracy = 0.1), ")"))

# Pivot to wide format - one row per year
emmeans_wide <- emmeans_df %>%
  select(marg_group, year, est_ci, prob) %>%
  pivot_wider(names_from  = marg_group,
              values_from = c(est_ci, prob)) %>%
  rename(less_ci   = `est_ci_Less marginalized`,
         more_ci   = `est_ci_More marginalized`,
         less_prob = `prob_Less marginalized`,
         more_prob = `prob_More marginalized`)

# Calculate gap in percentage points
emmeans_wide <- emmeans_wide %>%
  mutate(gap_pp = round(more_prob * 100, 1) - round(less_prob * 100, 1),
         Gap    = paste0("+", gap_pp, "pp"))

# Build gt table
emmeans_wide %>%
  select(year, less_ci, more_ci, Gap) %>%
  gt() %>%
  cols_label(
    year    = "Year",
    less_ci = "Less Marginalized",
    more_ci = "More Marginalized",
    Gap     = "Gap") %>%
  tab_header(
    title    = "Estimated Marginal Probabilities of Late Presentation",
    subtitle = "By year and marginalization group (mixed-effects logistic regression, nAGQ = 30)") %>%
  tab_spanner(
    label   = "Predicted Probability (95% CI)",
    columns = c(less_ci, more_ci)) %>%
  tab_style(
    style     = cell_fill(color = "#eef4fb"),
    locations = cells_body(columns = less_ci)) %>%
  tab_style(
    style     = cell_fill(color = "#fef1ef"),
    locations = cells_body(columns = more_ci)) %>%
  tab_style(
    style     = cell_text(color = "#c0392b", weight = "bold"),
    locations = cells_body(columns = Gap)) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  cols_align(align = "center", columns = c(less_ci, more_ci, Gap)) %>%
  cols_align(align = "left", columns = year) %>%
  gtsave("emmeans_table.png")
```

# Verify Model Fit

```{r}
# Compare all three models
AIC(model_glmer_main, model_glmer_linear, model_glmer)
BIC(model_glmer_main, model_glmer_linear, model_glmer)

# Joint test of interactions via likelihood ratio test
anova(model_glmer_main, model_glmer)
```

## Check random effects

```{r}
# Distribution of random effects - looks approximately normal
ranef_vals <- ranef(model_glmer)$munic_id[,1]

par(mfrow = c(1, 2))

qqnorm(ranef_vals, 
       main = "Q-Q plot of municipality random effects",
       cex.main = 0.85)
qqline(ranef_vals, col = "red")

hist(ranef_vals, breaks = 30,
     main = "Distribution of municipality random intercepts",
     xlab = "Random intercept value",
     col  = "steelblue",
     border = "white",
     cex.main = 0.85)

par(mfrow = c(1, 1))
```

## Check for influential municipios

```{r}
# Are any municipalities driving results?
ranef_df <- data.frame(munic_id = rownames(ranef(model_glmer)$munic_id),
                       intercept = ranef(model_glmer)$munic_id[,1])

# Flag potential outliers beyond 2 SDs
sd_re <- sd(ranef_df$intercept)
ranef_df %>%
  filter(abs(intercept) > 2 * sd_re) %>%
  arrange(desc(abs(intercept)))

# How many records are in these outlier municipalities?
outlier_munics <- ranef_df %>%
  filter(abs(intercept) > 2 * sd_re) %>%
  pull(munic_id)

SAEH_analytic %>%
  filter(munic_id %in% outlier_munics) %>%
  group_by(munic_id) %>%
  summarise(
    n = n(),
    prop_late = round(mean(late_pres), 3),
    marg_group = first(marg_group)) %>%
  arrange(desc(abs(prop_late - mean(SAEH_analytic$late_pres))))
```

# Layered plot

```{r}
# Recalculate observed proportions
observed_data <- SAEH_analytic %>%
  filter(!is.na(late_pres) & !is.na(marg_group)) %>%
  group_by(year, marg_group) %>%
  summarise(prop_late = mean(late_pres), .groups = "drop")

# Calculate observed means by group
obs_mean_less <- SAEH_analytic %>% 
  filter(marg_group == "Less marginalized") %>% 
  pull(late_pres) %>% mean(na.rm = TRUE)

obs_mean_more <- SAEH_analytic %>% 
  filter(marg_group == "More marginalized") %>% 
  pull(late_pres) %>% mean(na.rm = TRUE)

# Group-specific offsets
offset_less <- qlogis(obs_mean_less) - b0
offset_more <- qlogis(obs_mean_more) - (b0 + b_marg)

# Apply group-specific offsets
less_logits <- (b0 + offset_less) + b_yearc * year_range + b_yearsq * year_range^2
more_logits <- (b0 + offset_more + b_marg) +
               (b_yearc + b_int_c) * year_range +
               (b_yearsq + b_int_sq) * year_range^2

less_probs <- plogis(less_logits)
more_probs <- plogis(more_logits)

pred_data <- data.frame(
  year       = rep(actual_year, 2),
  prop_late  = c(less_probs, more_probs),
  marg_group = rep(c("Less marginalized", "More marginalized"),
                   each = length(actual_year)), 
  type = "Predicted")

# Plot
p3 <- ggplot() +
  geom_point(data = observed_data,
             aes(x = year, y = prop_late, color = marg_group),
             size = 2, shape = 16) +
  geom_line(data = observed_data,
            aes(x = year, y = prop_late, color = marg_group),
            linetype = "dashed", linewidth = 0.8) +
  geom_line(data = pred_data,
            aes(x = year, y = prop_late, color = marg_group),
            linewidth = 0.75) +
  scale_color_manual(values = c("Less marginalized" = "steelblue",
                                "More marginalized" = "tomato")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0.10, 0.30)) +
  scale_x_continuous(breaks = 2018:2024) +
  annotate("text", x = 2023.6, y = 0.120, label = "— Predicted",   size = 3.8, color = "black") +
  annotate("text", x = 2023.6, y = 0.105, label = "- - - Observed", size = 4,   color = "black") +
  labs(title = "Predicted vs. Observed Probability of Late Presentation by Year and Marginalization Group",
       x     = "Year of discharge",
       y     = "Probability of late presentation (≥13 weeks)",
       color = "Marginalization group") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11), 
        legend.position = "bottom")

p3
ggsave("layer_plot.png", p3, width = 9, height = 5, dpi = 300)
```

# Sensitivity Anaylsis with ordinal marginalization (quintiles):

```{r}
# Recode as ordered quintile factor
SAEH_analytic <- SAEH_analytic %>%
  mutate(grade_marg = factor(grade_marg,
                                  levels = c("Muy bajo", "Bajo", "Medio", "Alto", "Muy alto"),
                                  ordered = TRUE))

# Fit simple sensitivity model with ordinal marginalization
model_glmer_ord <- glmer(late_pres ~ grade_marg + year_c + year_sq + age_cat + 
                           parity_cat + indigena_bin + region + (1 | munic_id),
                         data    = SAEH_analytic,
                         family  = binomial(link = "logit"),
                         nAGQ    = 1,
                         control = glmerControl(optimizer = "bobyqa", 
                                                optCtrl  = list(maxfun = 2e5)))

summary(model_glmer_ord)

# Get predicted probabilities for each quintile
year_c_vals <- 2018:2024 - mean_year

# Get emmeans output
emmeans_ord <- emmeans(model_glmer_ord, ~ grade_marg * year_c,
                       at       = list(year_c = year_c_vals),
                       nuisance = c("age_cat", "parity_cat", "indigena_bin", "region"),
                       type     = "response")

# Convert to dataframe
emmeans_ord_df <- as.data.frame(emmeans_ord) %>%
  mutate(
    year   = round(year_c + mean_year),
    est_ci = paste0(
      scales::percent(prob, accuracy = 0.1),
      " (", scales::percent(asymp.LCL, accuracy = 0.1),
      "–", scales::percent(asymp.UCL, accuracy = 0.1), ")"),
    grade_marg = factor(grade_marg,
                        levels = c("Muy bajo", "Bajo", "Medio", "Alto", "Muy alto")))

# Pivot to wide - one row per year, one column per quintile
emmeans_ord_wide <- emmeans_ord_df %>%
  select(grade_marg, year, est_ci) %>%
  pivot_wider(names_from  = grade_marg,
              values_from = est_ci)

# Build gt table
emmeans_ord_wide %>%
  gt() %>%
  cols_label(
    year      = "Year",
    `Muy bajo` = "Muy Bajo\n(Q1)",
    `Bajo`     = "Bajo\n(Q2)",
    `Medio`    = "Medio\n(Q3)",
    `Alto`     = "Alto\n(Q4)",
    `Muy alto` = "Muy Alto\n(Q5)") %>%
  tab_header(
    title    = "Supplemental Table 2. Estimated Marginal Probabilities of Late Presentation by Marginalization Quintile and Year",
    subtitle = "Ordinal marginalization sensitivity analysis (mixed-effects logistic regression, nAGQ = 1)") %>%
  tab_source_note(
    "Values presented as predicted probability (95% CI), averaged over age, parity, indigenous identity, and region.
     Q1 = least marginalized (Muy bajo); Q5 = most marginalized (Muy alto).
     Linear contrast β_L = 0.226, p < 0.001; no significant higher-order terms.") %>%
  tab_spanner(
    label   = "Marginalization Quintile — Predicted Probability (95% CI)",
    columns = c(`Muy bajo`, `Bajo`, `Medio`, `Alto`, `Muy alto`)) %>%
  tab_style(
    style     = cell_fill(color = "#eef4fb"),
    locations = cells_body(columns = c(`Muy bajo`, `Bajo`))) %>%
  tab_style(
    style     = cell_fill(color = "#fef1ef"),
    locations = cells_body(columns = c(`Alto`, `Muy alto`))) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  cols_align(align = "center", 
             columns = c(`Muy bajo`, `Bajo`, `Medio`, `Alto`, `Muy alto`)) %>%
  cols_align(align = "left", columns = year) %>%
  gtsave("supplemental_table2_ordinal.png")
```

**every step up the marginalization scale corresponds to a higher probability of late presentation**

# Sensitivity Analysis for missing data (treat missing as a category):

```{r}
# Create missingness indicator in FULL pre-filtered dataset
SAEH_missing <- SAEH_full %>% 
  mutate(ga_missing = is.na(gestat),  # TRUE if gestational duration is missing
         late_pres = as.integer(gest_age_cat == "Late")) %>% # 1 = Late, 0 = Early
  arrange(munic_id, year)

SAEH_missing <- SAEH_missing %>% 
  mutate(late_pres_sens = case_when(late_pres == 1 ~ 1,
                                    late_pres == 0 ~ 0,
                                    is.na(gestat) ~ 2))  # missing as its own category

# Check if missingness is associated with marginalization
table(SAEH_missing$marg_group, SAEH_missing$ga_missing) %>% 
  prop.table(margin = 1) %>% 
  round(3)

# Chi-square test
chisq.test(SAEH_missing$marg_group, SAEH_missing$ga_missing)

# Check if missingness varies by year and marginalization
SAEH_missing %>%
  group_by(year, marg_group) %>%
  summarise(n_total   = n(),
            n_missing = sum(ga_missing),
            pct_missing = mean(ga_missing) * 100,
            .groups = "drop") %>%
  print(n = 50)

# Visualize missingness pattern
ggplot(SAEH_missing %>% 
         group_by(year, marg_group) %>%
         summarise(pct_missing = mean(ga_missing) * 100, .groups = "drop"),
       aes(x = year, y = pct_missing, color = marg_group, group = marg_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Less marginalized" = "steelblue",
                                "More marginalized" = "tomato")) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(title = "Percentage of missing gestational duration by year and marginalization group",
       x     = "Year",
       y     = "% missing gestational duration",
       color = "Marginalization group") +
  theme_minimal() +
  theme(legend.position = "bottom")


# Create centered year variable
mean_year <- mean(SAEH_missing$year, na.rm = TRUE)

# Now a year squared
SAEH_missing <- SAEH_missing %>%
  mutate(year_c  = year - mean_year,
         year_sq = year_c * year_c)

# Verify
SAEH_missing %>%
  select(year, year_c, year_sq) %>%
  distinct() %>%
  arrange(year)

# Logistic regression — does marginalization predict missingness?
model_missing <- glmer(ga_missing ~ marg_group * year_c + marg_group * year_sq + (1 | munic_id),
                       data    = SAEH_missing,
                       family  = binomial(link = "logit"),
                       nAGQ    = 1,
                       control = glmerControl(optimizer = "bobyqa",
                                              optCtrl  = list(maxfun = 2e5)))

summary(model_missing)
```

# Random slopes

```{r}
# Random intercept + random slope for linear time
model_glmer_rs <- glmer(
  late_pres ~ marg_group * year_c + marg_group * year_sq +
    age_cat + parity_cat + indigena_bin + region +
    (1 + year_c | munic_id),  # random slope for linear time
  data    = SAEH_analytic,
  family  = binomial(link = "logit"),
  nAGQ    = 1,  # start with 1 for speed
  control = glmerControl(optimizer = "bobyqa",
                         optCtrl  = list(maxfun = 2e5)))

# Compare to random intercept only model
anova(model_glmer, model_glmer_rs)
AIC(model_glmer, model_glmer_rs)
BIC(model_glmer, model_glmer_rs)
```

```{r}
# Random intercept + random slope using GLMMadaptive (correct approach)
model_rs <- mixed_model(
  fixed  = late_pres ~ marg_group * year_c + marg_group * year_sq +
             age_cat + parity_cat + indigena_bin + region,
  random = ~ year_c | munic_id,   # random slope for linear time
  data   = SAEH_analytic,
  family = binomial,
  nAGQ   = 20)   # start with 20, update to 40 if it converges

summary(model_rs)

# Update to 40 quadrature points for better accuracy
# model_rs40 <- update(model_rs, nAGQ = 40)
# summary(model_rs40)

# Note: anova() comparison between glmer and mixed_model objects
# won't work directly — compare AIC/BIC manually
AIC(model_glmer)   # your primary model
AIC(model_rs)    # random slopes model

BIC(model_glmer)
BIC(model_rs)
```
